Towards Explainability of Machine Learning Models in Insurance Pricing

### 2019-10-25 manuscript/manuscript.Rmd
some edits

# The Need to Understand the Black Box {#need}

@@

In this section, we attempt to develop a working definition of interpretability for ratemaking applications. While we will not provide a comprehensive survey of the prolific and fast evolving ML interpretability literature, we draw from it as appropriate in setting the stage for our discussion. Even among researchers in ML, there is not a consensus on the definition of interpretability; here are a few from frequently cited papers:

@@

Let us look at some examples of models that are considered by many to be interpretable. The GLM is probably the most oft-cited example of such a model (or rather, modeling framework) in the actuarial literature. Given a set of inputs, we can easily reason about what the output of the model is. As an illustrative example, consider a claim severity model with driver age, sex, and vehicle age as features; assuming a log link function and letting $Y$ denotes the response, we have

@@

Here, we can tell, for example, what the model would predict for the expected severity if we were to increase age by a certain amount, *all else being equal*, because the relationship between the feature and the response is simply multiplication by the coefficient $\beta_1$ and applying the inverse link function. Another commonly cited example of an interpretable model is a decision tree. An illustrative example is shown in Figure \@ref(fig:tree-plot1). Here, the prediction is arrived at by following a sequence of if-else decisions.

@@

Now, it is worth pointing out that, when declaring that GLM or decision trees are interpretable models, we are implicitly assuming that we are considering only a handful  of features. In fact, the ease at which we can reason about a model rapidly declines as the number of features, transformations, and interactions increase, as in the following (somewhat pathological) example:

@@

Similarly, one can see that in Figure \@ref(fig:tree-plot2), large and deep trees are tough to reason about. In other words, even when working within the framework of an "interpretable" class of models, we may still end up with something that many would consider "black box."

@@

Another occasional misconception is that we have no visibility into how some ML models *compute* predictions, which render them uninterpretable. Outside of proprietary algorithms, all common ML models, including neural networks, gradient boosted trees, and random forests, are well studied and have large bodies of literature documenting their inner workings. As an example, a fitted neural network model on tabular data is simply a composition of linear transformations followed with nonlinear activation functions. As in Equation \ref{eq:badglm}, one can write down the mathematical equation for calculating the prediction given some inputs, but it may be difficult for a human to reason about it. We show later that we can still provide explanations of completely "black box" models, but is important to note that ML model predictions are still governed by mathematical rules.

>>>>

# The Need to See Inside the Black Box {#need}

@@

In this section, we attempt to develop a working definition of interpretability for ratemaking applications. While we will not provide a comprehensive survey of the prolific and fast evolving ML interpretability literature, we draw from it as appropriate in setting the stage for our discussion. Even among researchers in the subject, there is not a consensus on the definition of interpretability; here are a few from frequently cited papers:

@@

In the actuarial science literature, the GLM is probably the most oft-cited example of an easily interpretable model. Given a set of inputs, we can easily reason about what the output of the model is. As an illustrative example, consider a claim severity model with driver age, sex, and vehicle age as features; assuming a log link function and letting $Y$ denotes the response, we have

@@

Here, we can tell, for example, what the model would predict for the expected severity if we were to increase age by a certain amount, *all else being equal*, because the relationship between the feature and the response is simply multiplication by the coefficient $\beta_1$ and applying the inverse link function.

Another commonly cited example of an interpretable model is a decision tree. An illustrative example is shown in Figure \@ref(fig:tree-plot1). Here, the prediction is arrived at by following a sequence of if-else decisions.

@@

Now, it is worth pointing out that, when declaring that GLMs or decision trees are interpretable models, we are implicitly assuming that we are considering only a handful  of features. In fact, the ease at which we can reason about a model declines as the number of features, transformations of them, and interactions increase, as in the following (somewhat pathological) example:

@@

Similarly, one can see that in Figure \@ref(fig:tree-plot2), larger trees are tough to reason about. In other words, even when working within the framework of an "interpretable" class of models, we may still end up with something that many would consider "black box."

@@

Another occasional misconception is that we have no visibility into how some ML models *compute* predictions, which renders them uninterpretable. Outside of proprietary algorithms, all common ML models, including neural networks, gradient boosted trees, and random forests, are well studied and have large bodies of literature documenting their inner workings. As an example, a fitted feedforward neural network is simply a composition of linear transformations followed by nonlinear activation functions. As in Equation \ref{eq:badglm}, one can write down the mathematical equation for calculating the prediction given some inputs, but it may be difficult for a human to reason about it. We show later that we can still provide explanations of completely "black box" models, but is important to note that ML model predictions are still governed by mathematical rules, and are deterministic in most cases.
